---
title: "chapter4"
output: html_notebook
---
#Chapter 4 The tidyverse

Up to now we have been manipulating vectors by reordering and subsetting them through indexing. However, once we start more advanced analyses, the preferred unit for data storage is not the vector but the data frame. In this chapter we learn to work directly with data frames, which greatly facilitate the organization of information. We will be using data frames for the majority of this book. We will focus on a specific data format referred to as tidy and on specific collection of packages that are particularly helpful for working with tidy data referred to as the tidyverse.

We can load all the tidyverse packages at once by installing and loading the tidyverse package:
```{r}
library(tidyverse)
```
We will learn how to implement the tidyverse approach throughout the book, but before delving into the details, in this chapter we introduce some of the most widely used tidyverse functionality, starting with the dplyr package for manipulating data frames and the purrr package for working with functions. Note that the tidyverse also includes a graphing package, ggplot2, which we introduce later in Chapter 7 in the Data Visualization part of the book; the readr package discussed in Chapter 5; and many others. In this chapter, we first introduce the concept of tidy data and then demonstrate how we use the tidyverse to work with data frames in this format.
##4.1 Tidy data
We say that a data table is in tidy format if each row represents one observation and columns represent the different variables available for each of these observations. The murders dataset is an example of a tidy data frame.
```{r}
#>        state abb region population total
#> 1    Alabama  AL  South    4779736   135
#> 2     Alaska  AK   West     710231    19
#> 3    Arizona  AZ   West    6392017   232
#> 4   Arkansas  AR  South    2915918    93
#> 5 California  CA   West   37253956  1257
#> 6   Colorado  CO   West    5029196    65
```

Each row represent a state with each of the five columns providing a different variable related to these states: name, abbreviation, region, population, and total murders.

To see how the same information can be provided in different formats, consider the following example:

```{r}
#>       country year fertility
#> 1     Germany 1960      2.41
#> 2 South Korea 1960      6.16
#> 3     Germany 1961      2.44
#> 4 South Korea 1961      5.99
#> 5     Germany 1962      2.47
#> 6 South Korea 1962      5.79
```
This tidy dataset provides fertility rates for two countries across the years. This is a tidy dataset because each row presents one observation with the three variables being country, year, and fertility rate. However, this dataset originally came in another format and was reshaped for the dslabs package. Originally, the data was in the following format:
```{r}
#>       country 1960 1961 1962
#> 1     Germany 2.41 2.44 2.47
#> 2 South Korea 6.16 5.99 5.79
```

The same information is provided, but there are two important differences in the format: 1) each row includes several observations and 2) one of the variables, year, is stored in the header. For the tidyverse packages to be optimally used, data need to be reshaped into tidy format, which you will learn to do in the Data Wrangling part of the book. Until then, we will use example datasets that are already in tidy format.

Although not immediately obvious, as you go through the book you will start to appreciate the advantages of working in a framework in which functions use tidy formats for both inputs and outputs. You will see how this permits the data analyst to focus on more important aspects of the analysis rather than the format of the data.

##4.2 Exercises
1. Examine the built-in dataset co2. Which of the following is true:

co2 is tidy data: it has one year for each row.
co2 is not tidy: we need at least one column with a character vector.
co2 is not tidy: it is a matrix instead of a data frame.
co2 is not tidy: to be tidy we would have to wrangle it to have three columns (year, month and value), then each co2 observation would have a row.
```{r}
datasets::co2
str(datasets::co2)
```
a

2. Examine the built-in dataset ChickWeight. Which of the following is true:

```{r}
datasets::ChickWeight
```

ChickWeight is not tidy: each chick has more than one row.
ChickWeight is tidy: each observation (a weight) is represented by one row. The chick from which this measurement came is one of the variables.
ChickWeight is not tidy: we are missing the year column.
ChickWeight is tidy: it is stored in a data frame.

b

3. Examine the built-in dataset BOD. Which of the following is true:
```{r}
datasets::BOD
```


BOD is not tidy: it only has six rows.
BOD is not tidy: the first column is just an index.
BOD is tidy: each row is an observation with two values (time and demand)
BOD is tidy: all small datasets are tidy by definition.
4. Which of the following built-in datasets is tidy (you can pick more than one):
c

BJsales
EuStockMarkets
DNase
Formaldehyde
Orange
UCBAdmissions
```{r}
datasets::BJsales

datasets::DNase
datasets::Formaldehyde
datasets::Orange
```
c,d,e

##4.3 Manipulating data frames
The dplyr package from the tidyverse introduces functions that perform some of the most common operations when working with data frames and uses names for these functions that are relatively easy to remember. For instance, to change the data table by adding a new column, we use mutate. To filter the data table to a subset of rows, we use filter. Finally, to subset the data by selecting specific columns, we use select.

###4.3.1 Adding a column with mutate
We want all the necessary information for our analysis to be included in the data table. So the first task is to add the murder rates to our murders data frame. The function mutate takes the data frame as a first argument and the name and values of the variable as a second argument using the convention name = values. So, to add murder rates, we use:


