---
title: "chapter4"
output: html_notebook
---
#Chapter 4 The tidyverse

Up to now we have been manipulating vectors by reordering and subsetting them through indexing. However, once we start more advanced analyses, the preferred unit for data storage is not the vector but the data frame. In this chapter we learn to work directly with data frames, which greatly facilitate the organization of information. We will be using data frames for the majority of this book. We will focus on a specific data format referred to as tidy and on specific collection of packages that are particularly helpful for working with tidy data referred to as the tidyverse.

We can load all the tidyverse packages at once by installing and loading the tidyverse package:
```{r}
library(tidyverse)
```
We will learn how to implement the tidyverse approach throughout the book, but before delving into the details, in this chapter we introduce some of the most widely used tidyverse functionality, starting with the dplyr package for manipulating data frames and the purrr package for working with functions. Note that the tidyverse also includes a graphing package, ggplot2, which we introduce later in Chapter 7 in the Data Visualization part of the book; the readr package discussed in Chapter 5; and many others. In this chapter, we first introduce the concept of tidy data and then demonstrate how we use the tidyverse to work with data frames in this format.
##4.1 Tidy data
We say that a data table is in tidy format if each row represents one observation and columns represent the different variables available for each of these observations. The murders dataset is an example of a tidy data frame.
```{r}
#>        state abb region population total
#> 1    Alabama  AL  South    4779736   135
#> 2     Alaska  AK   West     710231    19
#> 3    Arizona  AZ   West    6392017   232
#> 4   Arkansas  AR  South    2915918    93
#> 5 California  CA   West   37253956  1257
#> 6   Colorado  CO   West    5029196    65
```

Each row represent a state with each of the five columns providing a different variable related to these states: name, abbreviation, region, population, and total murders.

To see how the same information can be provided in different formats, consider the following example:

```{r}
#>       country year fertility
#> 1     Germany 1960      2.41
#> 2 South Korea 1960      6.16
#> 3     Germany 1961      2.44
#> 4 South Korea 1961      5.99
#> 5     Germany 1962      2.47
#> 6 South Korea 1962      5.79
```
This tidy dataset provides fertility rates for two countries across the years. This is a tidy dataset because each row presents one observation with the three variables being country, year, and fertility rate. However, this dataset originally came in another format and was reshaped for the dslabs package. Originally, the data was in the following format:
```{r}
#>       country 1960 1961 1962
#> 1     Germany 2.41 2.44 2.47
#> 2 South Korea 6.16 5.99 5.79
```

The same information is provided, but there are two important differences in the format: 1) each row includes several observations and 2) one of the variables, year, is stored in the header. For the tidyverse packages to be optimally used, data need to be reshaped into tidy format, which you will learn to do in the Data Wrangling part of the book. Until then, we will use example datasets that are already in tidy format.

Although not immediately obvious, as you go through the book you will start to appreciate the advantages of working in a framework in which functions use tidy formats for both inputs and outputs. You will see how this permits the data analyst to focus on more important aspects of the analysis rather than the format of the data.

##4.2 Exercises
1. Examine the built-in dataset co2. Which of the following is true:

co2 is tidy data: it has one year for each row.
co2 is not tidy: we need at least one column with a character vector.
co2 is not tidy: it is a matrix instead of a data frame.
co2 is not tidy: to be tidy we would have to wrangle it to have three columns (year, month and value), then each co2 observation would have a row.
```{r}
datasets::co2
str(datasets::co2)
```
a

2. Examine the built-in dataset ChickWeight. Which of the following is true:

```{r}
datasets::ChickWeight
```

ChickWeight is not tidy: each chick has more than one row.
ChickWeight is tidy: each observation (a weight) is represented by one row. The chick from which this measurement came is one of the variables.
ChickWeight is not tidy: we are missing the year column.
ChickWeight is tidy: it is stored in a data frame.

b

3. Examine the built-in dataset BOD. Which of the following is true:
```{r}
datasets::BOD
```


BOD is not tidy: it only has six rows.
BOD is not tidy: the first column is just an index.
BOD is tidy: each row is an observation with two values (time and demand)
BOD is tidy: all small datasets are tidy by definition.
4. Which of the following built-in datasets is tidy (you can pick more than one):
c

BJsales
EuStockMarkets
DNase
Formaldehyde
Orange
UCBAdmissions
```{r}
datasets::BJsales

datasets::DNase
datasets::Formaldehyde
datasets::Orange
```
c,d,e

##4.3 Manipulating data frames
The dplyr package from the tidyverse introduces functions that perform some of the most common operations when working with data frames and uses names for these functions that are relatively easy to remember. For instance, to change the data table by adding a new column, we use mutate. To filter the data table to a subset of rows, we use filter. Finally, to subset the data by selecting specific columns, we use select.

###4.3.1 Adding a column with mutate
We want all the necessary information for our analysis to be included in the data table. So the first task is to add the murder rates to our murders data frame. The function mutate takes the data frame as a first argument and the name and values of the variable as a second argument using the convention name = values. So, to add murder rates, we use:

```{r}
library(dplyr)
library(dslabs)
data("murders")
murders <- mutate(murders, rate = total / population * 100000)
murders
```
Notice that here we used total and population inside the function, which are objects that are not defined in our workspace. But why don’t we get an error?

This is one of dplyr’s main features. Functions in this package, such as mutate, know to look for variables in the data frame provided in the first argument. In the call to mutate above, total will have the values in murders$total. This approach makes the code much more readable.

We can see that the new column is added:
```{r}
head(murders)

```

Although we have overwritten the original murders object, this does not change the object that loaded with data(murders). If we load the murders data again, the original will overwrite our mutated version.

###4.3.2 Subsetting with filter
Now suppose that we want to filter the data table to only show the entries for which the murder rate is lower than 0.71. To do this we use the filter function, which takes the data table as the first argument and then the conditional statement as the second. Like mutate, we can use the unquoted variable names from murders inside the function and it will know we mean the columns and not objects in the workspace.
```{r}
filter(murders, rate <= 0.71)
```
4.3.3 Selecting columns with select
Although our data table only has six columns, some data tables include hundreds. If we want to view just a few, we can use the dplyr select function. In the code below we select three columns, assign this to a new object and then filter the new object:
```{r}
new_table <- select(murders, state, region, rate)
filter(new_table, rate <= 0.71)

```
4.3.3 Selecting columns with select
Although our data table only has six columns, some data tables include hundreds. If we want to view just a few, we can use the dplyr select function. In the code below we select three columns, assign this to a new object and then filter the new object:
```{r}
new_table <- select(murders, state, region, rate)
filter(new_table, rate <= 0.71)
```

In the call to select, the first argument murders is an object, but state, region, and rate are variable names.


4.4 Exercises
1. Load the dplyr package and the murders dataset.
```{r}
library(dplyr)
library(dslabs)
data(murders)
```
You can add columns using the dplyr function mutate. This function is aware of the column names and inside the function you can call them unquoted:
```{r}
murders <- mutate(murders, population_in_millions = population / 10^6)
```

We can write population rather than murders$population. The function mutate knows we are grabbing columns from murders.

Use the function mutate to add a murders column named rate with the per 100,000 murder rate as in the example code above. Make sure you redefine murders as done in the example code above ( murders <- [your code]) so we can keep using this variable.
```{r}
mutate(murders, rate =  murders$total/murders$population_in_millions)
```



2. If rank(x) gives you the ranks of x from lowest to highest, rank(-x) gives you the ranks from highest to lowest. Use the function mutate to add a column rank containing the rank, from highest to lowest murder rate. Make sure you redefine murders so we can keep using this variable.

```{r}
data(murders)
murder_rate <- murders$total/murders$population*100000
rate <- murder_rate

#*****
murders<- mutate(murders, rate= murder_rate, rank=rank(-rate))
murders
```


3. With dplyr, we can use select to show only certain columns. For example, with this code we would only show the states and population sizes:
```{r}
select(murders, state, abb)
```




4. The dplyr function filter is used to choose specific rows of the data frame to keep. Unlike select which is for columns, filter is for rows. For example, you can show just the New York row like this:
```{r}
filter(murders, state == "New York")
```

You can use other logical vectors to filter rows.

Use filter to show the top 5 states with the highest murder rates. After we add murder rate and rank, do not change the murders dataset, just show the result. Remember that you can filter based on the rank column.

```{r}

```

5. We can remove rows using the != operator. For example, to remove Florida, we would do this:

```{r}
no_florida <- filter(murders, state != "Florida")

no_south <- filter(murders, region != "South")
no_south
nrow(no_south)
```

6. We can also use %in% to filter with dplyr. You can therefore see the data from New York and Texas like this:
```{r}
filter(murders, state %in% c("New York", "Texas"))
murders_nw <- filter(murders, region %in% c('Northeast', 'West'))
murders_nw
nrow(murders_nw)
```
7. Suppose you want to live in the Northeast or West and want the murder rate to be less than 1. We want to see the data for the states satisfying these options. Note that you can use logical operators with filter. Here is an example in which we filter to keep only small states in the Northeast region.
```{r}
filter(murders, population < 5000000 & region == "Northeast")
murders<-mutate(murders, rate = murder_rate)

my_states <- filter(murders, region %in% c('Northeast', 'West')& rate < 1)
select(my_states, state, rate, rank)

```


Make sure murders has been defined with rate and rank and still has all states. Create a table called my_states that contains rows for states satisfying both the conditions: it is in the Northeast or West and the murder rate is less than 1. Use select to show only the state name, the rate, and the rank.



